Hospitalizations for psychiatric conditions among adolescents during the COVID-19 pandemic: adapted to CHOP input format

### 2021-12-15 psy_peds_3_CHOP.Rmd
adding the adapted code to incorporate new site results

---
title: "Hospitalizations for psychiatric conditions among adolescents during the COVID-19 pandemic"
output:
  html_document:
    code_folding: hide
    toc: yes
    toc_float: yes
editor_options: 
  markdown: 
    wrap: 72
---

@@

```{r message=FALSE, warning=FALSE}
folder_4ce_files <- "Input/09212021_2/"
obfuscation =  FALSE
#obfuscation = 3
raceAvailable = TRUE
#raceAvailable = FALSE
dateFormat <- "%d-%b-%y"
min_hosp_days <- 2
data_update_date <- "2021-06-05"
time_period <-"months" #possible values: days, weeks, months
```

@@

We will use as input the 2.2 data files. Specifically:

- LocalPatientSummary

- LocalPatientObservation

- LocalPatientClinicalcourse

For sites recording the race, we will also use an additional file:

- LocalPatientRace

### Reading 4CE phase 2.2 files

```{r message=FALSE, warning=FALSE}
### Read the CSV input files
source("R/readInputFiles.R")

getwd()
files <- readInputFiles( path      = folder_4ce_files, 
                         separator = ",",
                         skip      = 0, 
                         verbose   = FALSE )

## Create the output folder if it doesn't exist
#if (! "output" %in% list.dirs()) dir.create("output")

### Extract the patient summary and observation information. 
demo_raw <- files[["patientSummary"]] %>% filter(cohort == "AllAdm")
obs_raw <- files[["patientObservations"]] %>% filter(cohort == "AllAdm")
clinical_raw <- files[["patientClinicalCourse"]] %>% filter(cohort == "AllAdm")

### Read the file containing race information for those sites recording this variable
if( raceAvailable == TRUE ){
  race_raw <- read.delim(file.path(folder_4ce_files, "/LocalPatientRace.csv"), sep = ",", skip = 0)
}
```

### Read the ICD10 codes file

The file containing the ICD10 psychiatric related codes is located in
the `public-data` folder of the GitHub repository.

```{r}
icdCodes <- read.csv("public-data/pediatric_psychiatric_ICD10_codesV2.csv", header = TRUE, colClasses = "character") %>%
   filter( flag %in% c("originalCodes", "FrenchSpecificCodes")) %>%
   dplyr::select( disorder_group, ICD10_Code, description )
```

## Data-management

### Filtering on "inpatient" status

```{r}
# Filtering on inpatients here (in_hospital == 1), to reduce subsequent compute time of the encounter length calculation 
clinical_raw <- filter(clinical_raw, in_hospital == 1)
```

### Calculating duration of individual hospitalizations

Adding columns with hospitalization number and hospitalization length to the clinical raw file

```{r}
count_sequences_hospitalisation <- function(df, ...) {
  seq_hospitalisation_df <- data.frame(total_span = seq(min(df$days_since_admission),
                                                        max(df$days_since_admission))
  ) %>%
    left_join(df, by = c("total_span" = "days_since_admission")) %>%
    replace_na(list(in_hospital = 0))
  count_sequences <- rle(seq_hospitalisation_df$in_hospital)
  count_sequences_1 <- lapply(count_sequences, function(x) x[count_sequences$values == 1])
  n_sequences <- seq_along(count_sequences_1$lengths)
  sequences <- rep.int(n_sequences, count_sequences_1$lengths)
  sequences_len <- rep.int(count_sequences_1$lengths, count_sequences_1$lengths)
  stopifnot(length(df$days_since_admission) == length(sequences))
  data.frame(days_since_admission = df$days_since_admission,
             n_hospitalisation = sequences,
             len_hospitalisation = sequences_len)
}
stopifnot(all(clinical_raw$in_hospital == 1))
hospitalisations_seq_df <- clinical_raw %>%
  distinct(patient_num, cohort, days_since_admission, in_hospital) %>%
  group_by(patient_num, cohort) %>%
  group_modify(count_sequences_hospitalisation)

clinical_raw <- left_join(clinical_raw,
                          hospitalisations_seq_df,
                          by = c("patient_num", "cohort", "days_since_admission"))
```

### Counting ICD10 mental health codes in the obs_raw table

```{r}
### check that the ICD codes follow the regular expression [A-Z][0-9][0-9AB]\.?[0-9A-TV-Z]{0,4}
codesToReview <- obs_raw %>% dplyr::filter(concept_type == "DIAG-ICD10",
                                           ! grepl( "[A-Z][0-9][0-9AB]\\.?[0-9A-TV-Z]{0,4}", concept_code))
print(codesToReview)

### Calculate how many of the mental codes are in the dataset
mentalHealthCodes <- unique( obs_raw %>%
  dplyr::filter( concept_type == "DIAG-ICD10", 
                 concept_code %in% icdCodes$ICD10_Code ) %>%
  dplyr::mutate( concept_code = as.character( concept_code ), 
                 concept_length = nchar( concept_code ), 
                 concept_check = ifelse( concept_length == 4 | concept_length > 8, "Check", "OK")) %>%
  dplyr::select( concept_code, concept_length, concept_check) )

summary(as.factor( mentalHealthCodes$concept_check ) )

print( paste0("There are ", length(mentalHealthCodes$concept_code), " ICD10 mental codes out of the total ",length(unique(icdCodes$ICD10_Code)), " in the dataset"))
```

### Calculating age at hospitalization time
```{r}
# Calculating patient age at the time of visit, assuming the age used is the patients' age at the time of extraction (last discharge date available in the data)
obs_raw_age <- left_join(obs_raw, demo_raw[c("patient_num", "age")], by = c("patient_num")) %>% 
  inner_join(clinical_raw[, c("patient_num", "days_since_admission", "calendar_date")], 
            by = c("patient_num", "days_since_admission"))
obs_raw_age$age_time_diagnosis <- obs_raw_age$age - floor(as.numeric(as.Date(data_update_date) - as.Date(obs_raw_age$calendar_date, dateFormat)) / 365.25)
```

### Applying inclusion criteria to filter and format the data

We are selecting hospitalizations based on the following criteria:

- Associated with at least 1 ICD code

- With a duration of hospitalization spanning over at least two consecutive days

- Starting on or after 2019/01/01

``` {r}
obs_filtered_age <- obs_raw_age %>% 
  filter(concept_type == 'DIAG-ICD10', 
         days_since_admission >= 0)

clinical_filtered <- clinical_raw %>%
  filter( len_hospitalisation >= min_hosp_days, 
          cohort  %in% c("AllAdm")) %>%
  dplyr::mutate( date = as.Date( calendar_date, format = dateFormat ),
                 weeks = as.Date(cut( date, breaks = "week")),
                 month = as.Date(cut( date, breaks = "month")),
                 year = format( date, "%Y"), 
                 period = ifelse( date < pandemic_start_date,
                                  "before_pandemic", "during_pandemic"))

if( time_period == "weeks"){
  clinical_filtered$time_p <- clinical_filtered$weeks
}else if(time_period == "days"){
  clinical_filtered$time_p <- clinical_filtered$date
}else if(time_period == "months"){
  clinical_filtered$time_p <- clinical_filtered$month
}

```{r}
# Steps necessary to filter rows:
  # - encounters without ICD
  # - ICD diagnosed outside of encounters
  # - patients from the patients table that don't have any encounters matching the criteria
clinical_table <- clinical_filtered %>% 
  inner_join( distinct(obs_filtered_age, patient_num, days_since_admission),  by = c("patient_num", "days_since_admission")) %>% 
  distinct(patient_num, n_hospitalisation) %>% 
  inner_join(clinical_filtered, by = c("patient_num", "n_hospitalisation"))
# Adding new columns
observation_table <- obs_filtered_age %>% 
  inner_join(clinical_table[, c("patient_num", "days_since_admission")], by = c("patient_num", "days_since_admission"))
demographic_table <- demo_raw %>% 
  inner_join(distinct(clinical_table, patient_num), by = "patient_num")
all_codes <- observation_table %>% 
   inner_join(select(clinical_table, -calendar_date), by = c("patient_num", "days_since_admission")) %>% 
   inner_join(demographic_table[, c("patient_num", "sex")], by = "patient_num")
```

```{r}
stopifnot(unique(all_codes$cohort) == "AllAdm")
stopifnot(unique(all_codes$in_hospital) == 1)
stopifnot(unique(all_codes$len_hospitalisation >= min_hosp_days))
stopifnot(nrow(all_codes) == nrow(observation_table))
stopifnot(length(unique(all_codes$patient_num)) == length(unique(demographic_table$patient_num)))
stopifnot(length(unique(all_codes$patient_num)) == length(unique(observation_table$patient_num)))
stopifnot(length(unique(all_codes$patient_num)) == length(unique(clinical_table$patient_num)))
stopifnot(nrow(distinct(all_codes, patient_num, n_hospitalisation)) == nrow(distinct(clinical_table, patient_num, n_hospitalisation)))
stopifnot(nrow(distinct(all_codes, patient_num, days_since_admission)) == nrow(distinct(observation_table, patient_num, days_since_admission)))
```

### Filtering on age 11-17

```{r}
# Filtering hospitalizations where patient age is in inclusion criteria
observation_table_ado <- filter(observation_table, age_time_diagnosis >= 11 & age_time_diagnosis <= 17)

# Filtering observation tables which patient_num and days_since_admission are still in clinical table after filtering on age

clinical_test <- clinical_table[, c("patient_num", "n_hospitalisation", "days_since_admission")] %>% 
  inner_join(distinct(observation_table_ado, patient_num, days_since_admission, age_time_diagnosis), 
            by = c("patient_num", "days_since_admission")) %>% 
  distinct(patient_num, n_hospitalisation, age_time_diagnosis)

clinical_table_ado <- clinical_table %>% 
  inner_join(clinical_test,
             by = c("patient_num", "n_hospitalisation"))

# Filtering observation tables which patient_num are still in clinical table after filtering on age
demographic_table_ado <- demographic_table %>% 
  inner_join(distinct(clinical_table_ado, patient_num), by = "patient_num")

all_codes_ado <- all_codes %>% 
   filter(age_time_diagnosis >= 11 & age_time_diagnosis <= 17)
```

```{r}
stopifnot(unique(all_codes_ado$cohort) == "AllAdm")
stopifnot(unique(all_codes_ado$in_hospital) == 1)
stopifnot(unique(all_codes_ado$len_hospitalisation >= min_hosp_days))
stopifnot(nrow(all_codes_ado) == nrow(observation_table_ado))
stopifnot(length(unique(all_codes_ado$patient_num)) == length(unique(demographic_table_ado$patient_num)))
stopifnot(length(unique(all_codes_ado$patient_num)) == length(unique(observation_table_ado$patient_num)))
stopifnot(length(unique(all_codes_ado$patient_num)) == length(unique(clinical_table_ado$patient_num)))
stopifnot(nrow(distinct(all_codes_ado, patient_num, n_hospitalisation)) == nrow(distinct(clinical_table_ado, patient_num, n_hospitalisation)))
stopifnot(nrow(distinct(all_codes_ado, patient_num, days_since_admission)) == nrow(distinct(observation_table_ado, patient_num, days_since_admission)))
```

#### QC step: do a summary of the mental health ICD codes present in each site
As output we will generate a table with: the number of observations per code, number of patients diagnosed with each code and the date range in which the code was used.

```{r}
#filter the whole dataset to only mental codes
mental_codes_qc <- all_codes_ado %>% 
  filter( concept_code %in% mentalHealthCodes$concept_code )

mental_codes_qc_output <- mental_codes_qc %>%
  group_by( concept_code ) %>%
  dplyr::summarise( distinct_patients = ifelse(length(unique(patient_num)) > obfuscation | isFALSE(obfuscation),
                               length(unique(patient_num)), 
                               0.5),
                    distinct_observations = ifelse(length(patient_num) > obfuscation | isFALSE(obfuscation),
                               length(patient_num), 
                               0.5),                           
                    min_date = min( dmy( calendar_date ) ), 
                    max_date = max( dmy( calendar_date ) ))%>% 
  arrange(desc(distinct_observations))

mental_codes_qc_output <- left_join( mental_codes_qc_output, icdCodes, by=c("concept_code"= "ICD10_Code")) 
mental_codes_qc_output  <- mental_codes_qc_output[, c("disorder_group", "description", "concept_code", "distinct_patients", 
                                                    "distinct_observations", "min_date", "max_date")]

mental_codes_qc_output[1:10, ]
```

### Add the psychiatric/mental health ICD code description and filter

```{r message=FALSE, warning=FALSE}
# Getting site name, will be used later in the analysis
site <- unique(as.character(demo_raw$siteid))

icdCategory <- read.delim("public-data/icd10Codes.txt",
                          header = FALSE,
                          colClasses = "character")
colnames(icdCategory) <- c("icd_code_letter", "icd_description_category")

all_codes_desc <- all_codes_ado %>% 
    left_join( icdCodes, by=c("concept_code" = "ICD10_Code") ) %>%
  mutate(icd_code_category = ifelse(concept_code %in% icdCodes$ICD10_Code, 
                               "psy", 
                               "others")) %>%
  dplyr::select( patient_num, sex, age_time_diagnosis, concept_type, concept_code, calendar_date,
                 in_hospital, n_hospitalisation, len_hospitalisation, date, weeks, month, year,
                 period, time_p, disorder_group, description, icd_code_category )

all_codes_desc <- all_codes_desc %>% 
  mutate( icd_code_letter = ifelse(substr(concept_code, 1, 1) %in% c("D", "H"), substr(concept_code, 1, 2), substr(concept_code, 1, 1) ))

all_codes_desc <- left_join( all_codes_desc, icdCategory) %>%
  mutate( icd_description_category = replace_na( ifelse( concept_code %in% icdCodes$ICD10_Code, "Psy", icd_description_category), "Others")) 
all_codes_psy <- all_codes_desc %>% 
  filter(concept_code %in% icdCodes$ICD10_Code)

if( raceAvailable == TRUE ){
  all_codes_psy <- all_codes_psy %>%
    left_join( race_raw, by = c("patient_num") )
}
```

@@

```{r}
new_hospitalizations <- clinical_table %>%
  group_by(patient_num, n_hospitalisation) %>%
  slice(which.min(time_p)) %>%
  filter( in_hospital == 1 ) %>%
    group_by( time_p ) %>%
  summarise(count = ifelse(n_distinct(patient_num) > obfuscation | isFALSE(obfuscation),
                               n_distinct(patient_num), 
                               0.5)) %>% 
  ungroup()

nh <- admission_plot_count(df = new_hospitalizations, 
                     x = "time_p",
                     y = "count", 
                     title = paste0("New admissions (per ", time_period,")"), 
                     y_title = "Patient Count")

nh
```

## New admissions age range 11-17

```{r}
new_hospitalizations_ado <- clinical_table_ado %>%
  group_by(patient_num, n_hospitalisation) %>%
  slice(which.min(time_p)) %>%
  filter( in_hospital == 1 ) %>%
    group_by( time_p ) %>%
  summarise(count = ifelse(n_distinct(patient_num) > obfuscation | isFALSE(obfuscation),
                               n_distinct(patient_num), 
                               0.5)) %>% 
  ungroup()

nh_ado <- admission_plot_count(df = new_hospitalizations_ado, 
                     x = "time_p",
                     y = "count", 
                     title = paste0("New admissions (per ", time_period, ")"), 
                     y_title = "Patient Count")+
    labs( subtitle = "(age range 11-17)")

nh_ado
```

@@

```{r}
unique_patients_hospitalized <- clinical_table_ado %>%
  filter(date < end_date_plots &
         date >= start_date_plots ) %>%
    group_by( time_p ) %>%
  summarise(count = ifelse(n_distinct(patient_num) > obfuscation | isFALSE(obfuscation),
                               n_distinct(patient_num), 
                               0.5))

th <- admission_plot_count(df = unique_patients_hospitalized, 
                     x = "time_p",
                     y = "count", 
                     title = paste0("Total hospitalization (per ", time_period, ")"), 
                     y_title = "Patient Count") +
  labs( subtitle = "(age range 11-17)")
th
```

@@

```{r}
count_icd_psy <- all_codes_psy %>%
  filter(date < end_date_plots &
         date >= start_date_plots ) %>%
    group_by( time_p ) %>%
  summarise(count_psy =  ifelse(n_distinct(patient_num) > obfuscation | isFALSE(obfuscation),
                               n_distinct(patient_num), 
                               0.5)) %>%
  mutate( period = ifelse( time_p < pandemic_start_date,
                           "before_pandemic",
                           "during_pandemic")
  )

count_icd <- all_codes_desc %>%
  filter(date < end_date_plots &
         date >= start_date_plots ) %>%
    group_by( time_p ) %>%
  summarise(count_icd =  ifelse(n_distinct(patient_num) > obfuscation | isFALSE(obfuscation),
                               n_distinct(patient_num), 
                               0.5)) %>%
  mutate( period = ifelse( time_p < pandemic_start_date,
                           "before_pandemic",
                           "during_pandemic")
  )

count_icd <- count_icd %>%
  left_join(count_icd_psy, 
            by = c("time_p", "period")) %>% 
  replace_na(list(count_psy = 0))

count_icd$percentage_psy <- 100*count_icd$count_psy / count_icd$count_icd

thpd_psyCount <- count_icd %>%
  pivot_longer(names_to = "counts", cols=c(count_psy, count_icd)) %>%
  mutate( counts = ifelse( counts == "count_psy", "Psy", "Total")) %>%
  ggplot(aes(x = time_p, y = value, fill = counts, color = counts)) +
  geom_point() +
  # geom_line() +
  geom_vline(xintercept = as.Date(pandemic_start_date),
             linetype = "dashed") +
  scale_fill_manual(values = cbPalette) +
  scale_color_manual(values = cbPalette) +
  labs(y = "Counts",
       x = paste0("Date (by ", time_period,")"),
       title = paste0("Per ", time_period , " patient counts with mental health related ICD codes")) + 
  geom_smooth(method = "lm", formula = y ~ splines::ns(x, 2), se = TRUE, aes(colour=period), colour= "black", size = 0.5)
thpd_psyCount

@@

```{r}
all_patients <- unique( all_codes_desc[, c("patient_num", "time_p")])
patients_with_psy_conditions <- unique( all_codes_psy[, c("patient_num", "time_p")] )
patients_without_psy_conditions <- setdiff(all_patients,patients_with_psy_conditions)

patients_without_psy_conditions_counts <- patients_without_psy_conditions %>%
  group_by( time_p ) %>%
  summarise(count_no_psy_patients =  ifelse(n_distinct(patient_num) > obfuscation | isFALSE(obfuscation),
                               n_distinct(patient_num), 
                               0.5)) %>%
  mutate( period = ifelse( time_p < pandemic_start_date,
                           "before_pandemic",
                           "during_pandemic"))

patients_with_psy_conditions_counts <-  patients_with_psy_conditions %>%
  group_by( time_p ) %>%
  summarise(count_psy_patients =  ifelse(n_distinct(patient_num) > obfuscation | isFALSE(obfuscation),
                               n_distinct(patient_num), 
                               0.5)) %>%
  mutate( period = ifelse( time_p < pandemic_start_date,
                           "before_pandemic",
                           "during_pandemic"))

ratios_patients_with_without_psy <- patients_without_psy_conditions_counts %>%
  full_join( patients_with_psy_conditions_counts, by=c("time_p", "period")) %>%
  mutate( ratio = count_psy_patients/count_no_psy_patients, 
          percentage = 100 * count_psy_patients/(count_no_psy_patients + count_psy_patients))

@@

## Psychiatric conditions aggregating by subcategories

### New admissions: count

```{r}
count_disorder_group <- all_codes_psy %>%
  filter(date < end_date_plots &
         date >= start_date_plots ) %>%
    group_by( time_p, disorder_group ) %>%
  summarise(count_dg =  ifelse(n_distinct(patient_num) > obfuscation | isFALSE(obfuscation),
                               n_distinct(patient_num), 
                               0.5)) %>%
  mutate( period = ifelse( time_p < pandemic_start_date,
                           "before_pandemic",
                           "during_pandemic")
  )

dg <- count_disorder_group %>%  
  ggplot(aes(x = time_p, y= count_dg, fill = period, group = disorder_group )) +
  geom_bar(stat = "identity")+
  geom_vline(xintercept = as.Date(pandemic_start_date),
             linetype = "dashed") +
  geom_smooth(method = "lm", formula = y ~ splines::ns(x, 2), se= TRUE, aes(colour=period), colour= "black", size = 0.5) + 
  labs(x = time_period,
       y = "Patient Count",
       title = paste0("Per ", time_period, " patient counts with mental health related ICD codes"), 
       subtitle = "Grouped by mental health ICD subcategories") + 
  ylim(0, NA) + 
  facet_wrap(. ~ disorder_group, scales = "free_y") +
  theme(strip.text.x = element_text(size = 4),
        axis.text.x = element_text(size = 5, angle = 90), 
        axis.text.y = element_text(size = 6), 
        axis.title.x = element_text(size = 10),
        axis.title.y = element_text(size = 10),
        title = element_text(size = 12),
        plot.subtitle = element_text(size = 11),
        legend.position = "none") +
  scale_fill_manual(values = cbPalette[-1])

dg
```

#### Subset: the most prevalent disorder group
```{r}
dg_subset <- count_disorder_group %>%
  filter( disorder_group %in% c("Anxiety Disorders", "Depressive Disorders", "Suicide or Self-Injury" )) %>%
  ggplot(aes(x = time_p, y= count_dg, fill = period, group = disorder_group )) +
  geom_bar(stat = "identity")+
  geom_vline(xintercept = as.Date(pandemic_start_date),
             linetype = "dashed") +
  geom_smooth(method = "lm", formula = y ~ splines::ns(x, 2), se= TRUE, aes(colour=period), colour= "black", size = 0.5) + 
  labs(x = time_period,
       y = "Patient Count",
       title = paste0("Per ", time_period, " patient counts with mental health related ICD codes"), 
       subtitle = "Grouped by mental health ICD subcategories") + 
  ylim(0, NA) + 
  facet_wrap(. ~ disorder_group, scales = "free_y") +
  theme(strip.text.x = element_text(size = 12),
        axis.text.x = element_text(size = 12, angle = 90), 
        axis.text.y = element_text(size = 11), 
        axis.title.x = element_text(size = 12),
        axis.title.y = element_text(size = 11),
        title = element_text(size = 12),
        plot.subtitle = element_text(size = 11),
        legend.position = "none") +
  scale_fill_manual(values = cbPalette[-1])

dg_subset
```

### New admissions: percentage

```{r}
# estimate the counts
count_icd_all <- all_codes_desc %>%
  filter(date < end_date_plots &
         date >= start_date_plots ) %>%
    group_by( time_p ) %>%
  summarise(count_icd =  ifelse(n_distinct(patient_num) > obfuscation | isFALSE(obfuscation),
                               n_distinct(patient_num), 
                               0.5)) %>%
  mutate( period = ifelse( time_p < pandemic_start_date,
                           "before_pandemic",
                           "during_pandemic")
  )

# join the total with the disorder group count to estimate the percentage
perc_disorder_group <- count_icd_all %>%
  left_join(count_disorder_group, 
            by = c("time_p", "period")) %>% 
  replace_na(list(count_dg = 0))

perc_disorder_group$percentage_dg <- 100*perc_disorder_group$count_dg / perc_disorder_group$count_icd

dgPerc <- perc_disorder_group %>%  
  ggplot(aes(x = time_p, y= percentage_dg, fill = period, group = disorder_group )) +
  geom_bar(stat = "identity")+
  geom_vline(xintercept = as.Date(pandemic_start_date),
             linetype = "dashed") +
  geom_smooth(method = "lm", formula = y ~ splines::ns(x, 2), se= TRUE, aes(colour=period), colour= "black", size = 0.5) + 
  labs(x = time_period,
       y = "Patient (%)",
       title = paste0("Percentage of patients per ", time_period,"with mental health related ICD codes"), 
       subtitle = "Grouped by mental health ICD subcategories") + 
  ylim(0, NA) + 
  facet_wrap(. ~ disorder_group, scales = "free_y") +
  theme(strip.text.x = element_text(size = 4),
        axis.text.x = element_text(size = 5, angle = 90), 
        axis.text.y = element_text(size = 6), 
        axis.title.x = element_text(size = 10),
        axis.title.y = element_text(size = 10),
        title = element_text(size = 12),
        plot.subtitle = element_text(size = 11),
        legend.position = "none") +
  scale_fill_manual(values = cbPalette[-1])

dgPerc
```
```{r}
dgPerc_subset <- perc_disorder_group %>%  
  filter( disorder_group %in% c("Anxiety Disorders", "Depressive Disorders", "Suicide or Self-Injury" )) %>%
  ggplot(aes(x = time_p, y= percentage_dg, fill = period, group = disorder_group )) +
  geom_bar(stat = "identity")+
  geom_vline(xintercept = as.Date(pandemic_start_date),
             linetype = "dashed") +
  geom_smooth(method = "lm", formula = y ~ splines::ns(x, 2), se= TRUE, aes(colour=period), colour= "black", size = 0.5) + 
  labs(x = time_period,
       y = "Patient (%)",
       title = paste0("Percentage of patients per ", time_period,"with mental health related ICD codes"), 
       subtitle = "Grouped by mental health ICD subcategories") + 
  ylim(0, NA) + 
  facet_wrap(. ~ disorder_group, scales = "free_y") +
  theme(strip.text.x = element_text(size = 12),
        axis.text.x = element_text(size = 12, angle = 90), 
        axis.text.y = element_text(size = 12), 
        axis.title.x = element_text(size = 12),
        axis.title.y = element_text(size = 12),
        title = element_text(size = 12),
        plot.subtitle = element_text(size = 11),
        legend.position = "none") +
  scale_fill_manual(values = cbPalette[-1])

dgPerc_subset
```

## Psychiatric conditions, dichotomized on patients' sex

### Count

```{r}
# counts per time period and sex
patients_hospitalized_agg_sex <- all_codes_psy %>%
  filter(date < end_date_plots &
         date >= start_date_plots & 
           sex %in% c("female", "male")) %>%
    group_by( time_p, sex ) %>%
  summarise(count_sex =  ifelse(n_distinct(patient_num) > obfuscation | isFALSE(obfuscation),
                               n_distinct(patient_num), 
                               0.5))

patients_hospitalized_agg_sex %>%
  ggplot(aes(x = time_p, y = count_sex, fill = sex)) +
  geom_bar(stat="identity", position = "dodge")+
  scale_fill_manual(values = cbPalette) +
  labs(y = "Counts",
       x = time_period,
       title = paste0("Patient counts per ", time_period, " aggregated by sex"),
       subtitle = "Per Sex") + 
  theme(legend.position = "bottom") + 
  geom_smooth(method = "lm", formula = y ~ splines::ns(x, 2), se = TRUE, aes(colour=sex), colour= "black", size = 0.5)
```

### Percentage

```{r}
##percentage
total_patients_psy <- all_codes_psy %>%
  filter(date < end_date_plots &
         date >= start_date_plots ) %>%
    group_by( time_p ) %>%
  summarise(count =  ifelse(n_distinct(patient_num) > obfuscation | isFALSE(obfuscation),
                               n_distinct(patient_num), 
                               0.5))

patients_hospitalized_agg_sex_perc <- total_patients_psy %>%
  left_join( patients_hospitalized_agg_sex, by = c("time_p")) %>% 
  replace_na(list(count_sex = 0))

patients_hospitalized_agg_sex_perc$percentage <- 100*patients_hospitalized_agg_sex_perc$count_sex/patients_hospitalized_agg_sex_perc$count

patients_hospitalized_agg_sex_perc %>%
  ggplot(aes(x = time_p, y = percentage, fill = sex)) +
  geom_bar(stat="identity", position = "dodge")+
  scale_fill_manual(values = cbPalette) +
  labs(y = "Percentage ",
       x = time_period,
       title = paste0("Patient (%) per ", time_period," aggregated by sex"),
       subtitle = "Per Sex") + 
  theme(legend.position = "bottom") + 
  geom_smooth(method = "lm", formula = y ~ splines::ns(x, 2), se = TRUE, aes(colour=sex), colour= "black", size = 0.5)
```

## Table 1

```{r}

t.test2 <- function(m1, m2, s1, s2, n1, n2) {
  # t.test computed from sample statistics, ie using mean, standard deviation and count value of the samples
  # useful in case we can only retrieve aggregated stats from the sites
  # ref: https://stats.stackexchange.com/questions/30394/how-to-perform-two-sample-t-tests-in-r-by-inputting-sample-statistics-rather-tha
  # and https://en.wikipedia.org/wiki/Student%27s_t-test?oldformat=true#Equal_or_unequal_sample_sizes,_unequal_variances_(sX1_%3E_2sX2_or_sX2_%3E_2sX1)
  se <- sqrt( (s1^2/n1) + (s2^2/n2) )
  # welch-satterthwaite df
  df <- ( (s1^2/n1 + s2^2/n2)^2 )/( (s1^2/n1)^2/(n1-1) + (s2^2/n2)^2/(n2-1) )
  t <- (m1-m2)/se
  dat <- c(m1-m2, se, t, 2*pt(-abs(t),df))
  names(dat) <- c("Difference of means", "Std Error", "t", "p-value")
  return(dat)
}

age_by_period <- all_codes_psy %>%
  distinct(patient_num, age_time_diagnosis, period) %>% 
  dplyr::group_by(period) %>%
  dplyr::summarise( mean = mean( age_time_diagnosis ),
                    sd = sd( age_time_diagnosis ),
                    count = n_distinct(patient_num))

age_test <- t.test2(m1 = age_by_period[[1, "mean"]],
                    m2 = age_by_period[[2, "mean"]],
                    s1 = age_by_period[[1, "sd"]],
                    s2 = age_by_period[[2, "sd"]],
                    n1 = age_by_period[[1, "count"]],
                    n2 = age_by_period[[2, "count"]]) %>%
  `[[`("p-value")
```

```{r}
sex <- all_codes_psy %>%
  dplyr::distinct(patient_num, sex) %>% 
  dplyr::group_by( sex ) %>%
  dplyr::summarise( count = n_distinct(patient_num) ) %>%
  dplyr::mutate( count = ifelse( count > obfuscation, count, obfuscation * 0.5),
                 total = sum( count ), percentage = round( count/total*100, 2)  )

sex_by_period <- all_codes_psy %>%
  dplyr::distinct(patient_num, period, sex) %>% 
  group_by(period, sex ) %>%
  summarise(count = n_distinct(patient_num)) %>%
  group_by(period, .add=TRUE) %>%
  mutate( count = ifelse( count > obfuscation, count, obfuscation * 0.5),
          percentage=round(count/sum( count )*100, 2)) %>%
  replace_na(list(male = 0, female = 0))

sex_test <- sex_by_period %>%
  select(-percentage) %>%
  pivot_wider(names_from = sex, values_from = count) %>%
  ungroup() %>%
  select(male, female) %>%
  chisq.test() %>%
  `[[`("p.value")
```

```{r}
summary_diagnosis <- all_codes_psy %>%
  dplyr::distinct( patient_num, description, period, weeks, concept_code ) %>%
  dplyr::group_by( concept_code, description, period ) %>%
  dplyr::count( patient_num ) %>%
  dplyr::tally(name = "count") %>%
  dplyr::mutate(count = ifelse( count > obfuscation, count, obfuscation * 0.5)) %>%
  dplyr::arrange( desc(count) ) %>%
  dplyr::ungroup()

summary_diagnosis_test <- summary_diagnosis %>%
  tidyr::pivot_wider(names_from = period,
                     values_from = count) %>%
  #select(late_pandemic, early_pandemic) %>%
  select(during_pandemic) %>%
  #tidyr::replace_na(list(late_pandemic = 0, early_pandemic = 0)) %>%
  tidyr::replace_na(list(during_pandemic = 0)) %>%
  chisq.test() %>%
  `[[`("p.value")

```{r}
summary_disorder_group <- all_codes_psy %>%
  dplyr::group_by( disorder_group ) %>%
  dplyr::count( patient_num ) %>%
  dplyr::tally(name = "count") %>%
  dplyr::mutate(count = ifelse( count > obfuscation, count, obfuscation * 0.5)) %>%
  dplyr::arrange( desc( count ))

summary_disorder_group_by_period <- all_codes_psy %>%
  dplyr::distinct( patient_num, disorder_group, period, weeks ) %>%
  dplyr::group_by( period, disorder_group ) %>%
  summarise(count = n_distinct(patient_num)) %>%
  group_by(period, .add=TRUE) %>%
  mutate(count = ifelse( count > obfuscation, count, obfuscation * 0.5),
         percentage = count/sum( count )*100)

summary_disorder_group_by_period_test <- summary_disorder_group_by_period %>%
  pivot_wider(id_cols = disorder_group, names_from = period, values_from = c(count, percentage)) %>%
  mutate(across(where(is.numeric), ~replace_na(.x, replace = 0))) %>%
  #select(count_early_pandemic, count_late_pandemic) %>%
  select(count_during_pandemic) %>%
  chisq.test() %>%
  `[[`("p.value")

```{r}
format_counts <- function(count, percentage) {
  glue::glue("{count} ({percentage} %)", count = count, percentage = percentage)
}

age_table1 <- age_by_period %>% 
  mutate(result = glue::glue("{mean} (sd: {sd})", mean = round(mean, 2), sd = round(sd, 2))) %>% 
  select(result, period) %>% 
  pivot_wider( names_from = period, values_from = result ) %>%
  mutate(name = "Age")

>>>>

---
title: "Hospitalizations for psychiatric conditions among adolescents during the COVID-19 pandemic: adapted to CHOP input format"
output:
  html_document:
    code_folding: hide
    toc: yes
    toc_float: yes
editor_options: 
  markdown: 
    wrap: 72
---

@@

```{r message=FALSE, warning=FALSE}
raceAvailable = TRUE
#raceAvailable = FALSE
#data_update_date <- "2021-06-05"
time_period <-"months" #possible values: days, weeks, months
```

@@

We will use as input the files from CHOP with their counts.

### Reading the files with total counts per month and patients with psychiatric conditions per month.

```{r message=FALSE, warning=FALSE}
outputDir <- "../output/CHOP/"
allCounts <- read.csv( file = paste0( outputDir, "All-Counts.csv"))
psyCounts <- read.csv( file = paste0( outputDir, "Psyc-Counts.csv"))

allCounts <- allCounts %>%
  mutate( time_p = ymd(paste( allCounts$yr1, allCounts$mo1, "01", sep="-")), 
          period = ifelse( time_p < pandemic_start_date,
                           "before_pandemic",
                           "during_pandemic"), 
          siteid = "CHOP") %>%
  select( time_p, count_all = ct1, period, siteid )

psyCounts <- psyCounts %>%
  mutate( time_p = ymd(paste( psyCounts$yr2, psyCounts$mo2, "01", sep="-")), 
          period = ifelse( time_p < pandemic_start_date,
                           "before_pandemic",
                           "during_pandemic"), 
          siteid = "CHOP" ) %>%
  select( time_p, count_psy_patients = ct2, period, siteid )
```

@@

@@

```{r}
th <- admission_plot_count(df = allCounts, 
                     x = "time_p",
                     y = "count_all", 
                     title = paste0("Total hospitalization (per ", time_period, ")"), 
                     y_title = "Patient Count") +
  labs( subtitle = "(age range 11-17)")
th
```

@@

```{r}
thpd_psyCount <- psyCounts %>%
  ggplot(aes(x = time_p, y = count_psy_patients )) +
  geom_point() +
  geom_vline(xintercept = as.Date(pandemic_start_date),
             linetype = "dashed") +
  scale_fill_manual(values = cbPalette) +
  scale_color_manual(values = cbPalette) +
  labs(y = "Counts",
       x = paste0("Date (by ", time_period,")"),
       title = paste0("Per ", time_period , " patient counts with mental health related ICD codes")) + 
  geom_smooth(method = "lm", formula = y ~ splines::ns(x, 2), se = TRUE, aes(colour=period), colour= "black", size = 0.5)
thpd_psyCount

@@

```{r}
ratios_patients_with_without_psy <- inner_join( allCounts, psyCounts )

ratios_patients_with_without_psy <- ratios_patients_with_without_psy %>%
  group_by( time_p, period, siteid ) %>%
  mutate( count_no_psy_patients = count_all - count_psy_patients, 
          ratio = round( count_psy_patients / count_no_psy_patients, 3), 
          percentage = round( 100*count_psy_patients / count_all, 1) ) %>%
  select( time_p, count_no_psy_patients, period, count_psy_patients, ratio, percentage, siteid )

@@
